Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-07-01 docs/paper.Rmd
write theoretical background (recent refinements - credit constraints)

>>>>

\subsubsection{Recent Refinements to the Q-Q Model}

Contemporary research has further refined the quantity–quality model by relaxing some of its initial simplifying assumptions. Three important extensions involve (1) capital market imperfections that constrain parents’ ability to invest in child quality, (2) intra-household conflict and bargaining between mothers and fathers over fertility choices, and (3) recognition that child quality is multi-dimensional, which extends beyond schooling to include health, nutrition, and other facets of human capital.

To begin, I examine how credit constraints can give rise to poverty traps. The canonical quantity–quality (Q–Q) framework assumes that parents can reallocate resources freely; borrowing against future earnings to finance schooling or health investments whenever the expected return is high. In practice, especially in low-income settings, credit markets function imperfectly: poor households typically cannot secure loans to cover children’s education or medical costs even when such investments would yield substantial future gains. This market failure magnifies the Q–Q trade-off. Becker, Lewis, and Willis–already noted by \citet{grawe2008quality} as emphasizing “resource limitations”–implicitly recognized this issue, but contemporary models make it explicit by imposing a borrowing constraint. Parents must fund childrearing from current income alone; they cannot collateralize a child’s future wages to pay today’s school fees. Consequently, when income is low, each additional birth directly reduces the attainable quality per child, potentially trapping families in a low-income–high-fertility equilibrium.

Formal models show that “the quality–quantity trade-off arises from a binding credit constraint that prevents parents from borrowing against future child income.” Empirical work supports this mechanism. \citet{kremer2002income} and \citet{de2003inequality} document that countries facing tighter liquidity constraints tend to display higher fertility and lower educational attainment, consistent with liquidity-constrained parents favoring quantity over quality. Cross-country evidence likewise indicates that where financial frictions are more severe, the negative correlation between fertility and schooling is stronger. Theoretically, introducing a borrowing limit can generate multiple steady states: one with low fertility and high investment when incomes suffice to cover quality costs, and another with high fertility and minimal investment when they do not. Policy instruments such as education subsidies or conditional cash transfers effectively relax these constraints, which nudge households toward the low-fertility, high-investment equilibrium. In short, incorporating credit market imperfections deepens the explanatory power of the Q-Q model: economic growth alone may not reduce fertility if households remain too cash-poor to afford schooling, whereas targeted quality-enhancing transfers can catalyze both demographic and human-capital transitions.

### 2025-07-01 docs/paper.tex
write theoretical background (recent refinements - credit constraints)

>>>>

\subsubsection{Recent Refinements to the Q-Q Model}

Contemporary research has further refined the quantity--quality model by
relaxing some of its initial simplifying assumptions. Three important
extensions involve (1) capital market imperfections that constrain
parents' ability to invest in child quality, (2) intra-household
conflict and bargaining between mothers and fathers over fertility
choices, and (3) recognition that child quality is multi-dimensional,
which extends beyond schooling to include health, nutrition, and other
facets of human capital.

To begin, I examine how credit constraints can give rise to poverty
traps. The canonical quantity--quality (Q--Q) framework assumes that
parents can reallocate resources freely; borrowing against future
earnings to finance schooling or health investments whenever the
expected return is high. In practice, especially in low-income settings,
credit markets function imperfectly: poor households typically cannot
secure loans to cover children's education or medical costs even when
such investments would yield substantial future gains. This market
failure magnifies the Q--Q trade-off. Becker, Lewis, and Willis--already
noted by \citet{grawe2008quality} as emphasizing ``resource
limitations''--implicitly recognized this issue, but contemporary models
make it explicit by imposing a borrowing constraint. Parents must fund
childrearing from current income alone; they cannot collateralize a
child's future wages to pay today's school fees. Consequently, when
income is low, each additional birth directly reduces the attainable
quality per child, potentially trapping families in a
low-income--high-fertility equilibrium.

Formal models show that ``the quality--quantity trade-off arises from a
binding credit constraint that prevents parents from borrowing against
future child income.'' Empirical work supports this mechanism.
\citet{kremer2002income} and \citet{de2003inequality} document that
countries facing tighter liquidity constraints tend to display higher
fertility and lower educational attainment, consistent with
liquidity-constrained parents favoring quantity over quality.
Cross-country evidence likewise indicates that where financial frictions
are more severe, the negative correlation between fertility and
schooling is stronger. Theoretically, introducing a borrowing limit can
generate multiple steady states: one with low fertility and high
investment when incomes suffice to cover quality costs, and another with
high fertility and minimal investment when they do not. Policy
instruments such as education subsidies or conditional cash transfers
effectively relax these constraints, which nudge households toward the
low-fertility, high-investment equilibrium. In short, incorporating
credit market imperfections deepens the explanatory power of the Q-Q
model: economic growth alone may not reduce fertility if households
remain too cash-poor to afford schooling, whereas targeted
quality-enhancing transfers can catalyze both demographic and
human-capital transitions.
